Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-04 paper.qmd
Updated the MNIST script

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
names(training_data_pbmc) <- paste0("PC", 1:50)

@@

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "Reproduction of plot published in Chen et al. (2023) showing a $2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. The (hyper-)parameter settings, beyond the defaults, are 30 nearest neighbors and minimum distance equal to 0.3. We use our model-in-the-data-space to assess whether this is an accurate representation of structure present in the high-dimensional data, or if it is misleading."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H
#| eval: false

umap_pbmc 
```

@@

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC"
)

@@

pbmc_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("a2")

@@

pbmc_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("a3")

@@

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled_best,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC"
)

@@

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "UMAP",
  col_start = "PC")

@@

pbmc_proj_umap_model1_best <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("b2")

@@

pbmc_proj_umap_model2_best <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("b3")

>>>>

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
names(training_data_pbmc) <- paste0("p", 1:50)

@@

@@

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "p"
)

@@

pbmc_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("a2")

@@

pbmc_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("a3")

@@

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled_best,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "p"
)

@@

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "UMAP",
  col_start = "p")

@@

pbmc_proj_umap_model1_best <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("b2")

@@

pbmc_proj_umap_model2_best <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-1, 1)) +
  ylim(c(-1, 1)) +
  interior_annotation("b3")
